On the surprising effectiveness of a simple matrix exponential derivative approximation, with application to global SARS-CoV-2

### 2023-09-14 manuscript/pnas_revision.tex
first pass done

The continuous-time Markov chain (CTMC) \cite{norris1998markov}
represents the gold-standard mathematical model for such evolution of
characters (e.g., nucleotides) within a fixed span of evolutionary
time. A CTMC defined over a discrete, $d$-element state space consists
of a row vector $\ppi_0$ whose individual components describes the
probability of inhabiting each of the possible states at time $t=0$,
as well as a $d\times d$ infinitesimal generator (or rate) matrix
$\QQ$ with non-negative off-diagonal elements $q_{ij}$, $i\neq j$, and
non-positive diagonal elements $q_{ii}=-\sum_j q_{ij}$. For any lag
$t\geq 0$, the matrix exponential
\cite{moler1978nineteen,moler2003nineteen} provides the Markov chain's
transition probability matrix
\begin{align}
    \P_t := e^{t \QQ} := \sum_{n=0}^\infty \frac{t^n\QQ^n}{n!}  \, ,
\end{align}
which has elements $[\P_t]_{ij}$ that dictate the probability of the
process jumping from state $i$ to state $j$ after time $t$. It is
straightforward to verify that $\P_t$ is a valid transition matrix,
having probability vectors for rows: if $\one$ and $\Zero$ are the
column vectors of ones and zeros, respectively, then $\QQ\one=\Zero$
and, therefore, $\P_t\one=\one$. The law of total probability then
provides the marginal probability of the process at any time
$t \geq 0$ as $\ppi_t=\ppi_0 e^{t\QQ}$.

@@

    these eigenvalue are all non-positive,

>>>>

The continuous-time Markov chain (CTMC) \cite{norris_1997}
represents the gold-standard mathematical model for such evolution of
characters (e.g., nucleotides) within a fixed span of evolutionary
time. A CTMC defined over a discrete, $d$-element state space consists
of a row vector $\ppi_0$ whose individual components describes the
probability of inhabiting each of the possible states at time $t=0$,
as well as a $d\times d$ infinitesimal generator (or rate) matrix
$\QQ$ with non-negative off-diagonal elements $q_{ij}$, $i\neq j$, and
non-positive diagonal elements $q_{ii}=-\sum_j q_{ij}$. For any lag
$t\geq 0$, the matrix exponential
\cite{moler1978nineteen,moler2003nineteen} provides the Markov chain's
transition probability matrix
\begin{align}
    \P_t := e^{t \QQ} := \sum_{n=0}^\infty \frac{t^n\QQ^n}{n!}  \, ,
\end{align}
which has elements $[\P_t]_{ij}$ that dictate the probability of the
process jumping from state $i$ to state $j$ after time $t$. It is
straightforward to verify that $\P_t$ is a valid transition matrix,
having probability vectors for rows: if $\one$ and $\Zero$ are the
column vectors of ones and zeros, respectively, then $\QQ\one=\Zero$
and, therefore, $\P_t\one=\one$. The law of total probability then
provides the marginal probability of the process at any time
$t \geq 0$ as $\ppi_t=\ppi_0 e^{t\QQ}$.

@@

    these eigenvalues are all non-positive,

### 2023-09-14 manuscript/revision_letter.tex
first pass done

@@

@@

@@

>>>>

\newcommand{\Y}{\mathbf{Y}}
\begin{reply}
	We would prefer to keep this section the same, but we do relate the general statement to Bayesian inference in a footnote:
	\begin{quote}
		In the Bayesian inference context, we are typically
		considering the posterior density function
		$\pi(\ttheta):=p(\ttheta|\Y)$ for the parameter
		$\ttheta\in\mathbb{R}^K$ given observed data $\Y$.
	\end{quote}
\end{reply}

@@

\begin{reply}
TBD	
\end{reply}

@@

\begin{reply}
Thanks for catching this!  We have amended accordingly. 
\end{reply}

@@

\begin{reply}
	Again, thanks for catching this.  We have updated the citation to reference the specific chapter \cite{norris_1997}.
\end{reply}

\begin{reply}
	Again, we wish to thank you for putting so much time and effort in to your review of our manuscript.  It is greatly appreciated.  We have addressed and implemented your suggestions to the best of our abilities and are confident that the manuscript is now much improved.
\end{reply}
